---
title: "R_tutorial4"
author: "Sunwoo Jeong"
date: "12/4/2019"
output: html_document
---

# The data \& the packages

It would take us too far afield to cover all the theoretical background for linear models and mixed effects models. Please consult the readings I've posted for a gentle introduction, as well as a variety of resources available on the web. The course being offered next semester will also provide an in-depth training. Today, we will focus on the practical aspects of fitting linear regression models, logistic regression models, and mixed effects models. Let us first call in the data -- along with the R packages needed for fitting regression models.

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)

## Install these packages if they aren't already installed
# install.packages("lme4")
# install.packages("lmerTest")

## Call them in
library(lme4)
library(lmerTest)

## The data
datc <- read.csv('asmt1-korean-stops.csv')
datv <- read.csv('asmt1-korean-vowels.csv')
```


# The vowel data

Say that we want to examine if height and speaker gender is a significant predictor or F0.
A simple linear model which tests this would be as follows.
(syntax: dependent variable ~ predictor(factor)1 + predictor(factor)2, ..., data=data)


```{r}
m1 <- lm(F0 ~ ht + sp_gender, data=datv)
summary(m1) 
```

Any other things you might want to include in the model? Try fitting models with F1, F2, and F3 as dependent variables as well.

```{r}
m2 <- lm(F0 ~ ht + sp_gender + wpos + vpos, data=datv)
summary(m2) 

m3 <- lm(F1 ~ ht + sp_gender + vtype, data=datv)
summary(m3)
```
A sample report: Speaker gender was a significant predictor of F0: male speakers elicited significantly lower F0 than female speakers (beta = -148, SE = 17, t = -8, p < .001).

Focusing on m2, discuss among groups how these results might be interpreted.

Based on the discussion (and also ideally, supported by visualizations), we can make informed decisions about adapting our model to include the following interaction:

```{r}
datv$vpos <- as.factor(datv$vpos)
m4 <- lm(F0 ~ ht + sp_gender + wpos * vpos, data=datv)
summary(m4) 
```

# The consonant data

Consult the last tutorial materials if needed, and create a new column, stop_type, for datc, which codes whether a given stop of the word is a lenis one or an aspirated one. Then fit a linear model with VOT as the main dependent variable, and v1F0m and stop_type as predictors (independent variables). Discuss among groups and interpret the results.

```{r}
head(datc)
summary(datc$stop)

datc2 <- datc %>%
  mutate(stop_type = ifelse(stop == "b" | stop == "d" | stop == "g", "len", "asp"))

datc2$stop_type <- as.factor(datc2$stop_type)

m5 <- lm(VOT ~ stop_type + v1F0m, data=datc2)
summary(m5)
```
# The Korean Ganong data

Logistic regression is the way to go when the dependent variable involves binary choices. Mixed effects models are often used (positing random slopes and intercepts) to explain away any pontential (unaccounted) random effects. Usually, items and participants are included as random effects.

```{r}
# First, I ran the codes from the previous tutorial, and saved the .csv as follows.
write.csv(datg2, file = "korean-ganong-cleaned.csv",row.names=FALSE)

# Let us now call in the data
datg <- read.csv('korean-ganong-cleaned.csv')
head(datg)

datg$subject_nr <- as.factor(datg$subject_nr)
datg$F0 <- as.factor(datg$F0)

# Simple logistic regression model

m6 <- glm(stop_response2 ~ F0 + word_status + subj_gender, data = datg, family = 'binomial')
summary(m6)

# Mixed effects logistic regression model
# Function for the mixed effects linear regression model would be: lmer

m7 <- glmer(stop_response2 ~ F0 + word_status + subj_gender + (1|subject_nr), data = datg, family = 'binomial')
summary(m7)
```
Discuss among groups the results.
Any online measures you would like to examine?
```{r}
m8 <- lm(log(response_time) ~ F0 + word_status + base_word, data = datg)
summary(m8)
```
