---
title: "R Tutorial II: Data visualization via ggplot"
author: "Sunwoo Jeong"
date: "11/20/2019"
output: html_document
---

## Acknowledgement

This tutorial is based on [Mike Frank's tutorial of Tidyverse](https://github.com/mcfrank/tidyverse-tutorial), which in turn is based on Hadley Wickham's [R for data scientists](http://r4ds.had.co.nz/). The latter resource contains extensive instructions on how to create effective visualizations of your experimental results, some of which we will cover today. Let's begin by calling in the data we will be working with (your assignments 1 and 3) and the necessary package (tidyverse, which includes ggplot2, which is commonly used for creating plots and visualizations in R).

```{r setup, include=FALSE}
## Call this in whenever you want to use associated packages such as ggplot2, tidyr, etc.
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)

## The data
# datv is now fixed
datv <- read.csv('asmt1-korean-vowels.csv')
datc <- read.csv('asmt1-korean-stops.csv')
datg <- read.csv('asmt3-korean-ganong.csv')
```

<!-- ----------------------------------------------------------------------- -->
# Goals and Introduction

By the end of this tutorial, you will know:

+ How to create plots and visualizations from data sets
+ How to create plots and visualizations from data summaries

Take a quick look at how each data set look like:

```{r}
head(datv)
head(datc)
head(datg)
```

# Visualizations from data sets

What we will be working with is `ggplot2`. It is a plotting package that easily takes advantage of tidy data. ggplots have two important parts (there are of course more):

+ `aes` - the aesthetic mapping, or which data variables get mapped to which visual variables (x, y, color, symbol, etc.)
+ `geom` - the plotting objects that represent the data (points, lines, shapes, etc.)

Let's try plottng the raw F0 data from assignment 1, with speaker height on the x-axis, F0 on the y-axis, and speaker gender represented as color coded dots.

## The vowel data from assignment 1

```{r}
datv %>%
  ggplot(aes(x = vtype, y = F0, col = sp_gender)) + 
  geom_point() 
```

Any thoughts on the data? Are they in line with the argument made in Johnson (2006)? Suggestions on other modes of visualizations?


The plot above made use of the following aesthetic mappings: x-axis, y-axis, and colors. If you want to plot additional factors, you can also use panels (`facet_grid`), as follows:

```{r}
datv %>%
  ggplot(aes(x = vtype, y = F1, col = sp_gender)) + 
  geom_point() +
  facet_grid(. ~ vpos)


datv %>%
  ggplot(aes(x = ht, y = F0, col = sp_gender)) + 
  geom_point() +
  facet_grid(vtype ~ vpos)
```

What do the plots above visualize? Notice any interesting observations/generalizations, and/or problems of the experiment design?


**Exercise 1**

From datv, create a plot which maps wpos on panels, F0 on y-axis, and vtype on x-axis. (You may choose to color code speaker gender or not)

```{r}

```

**Exercise 2**

From datv, create a plot which maps wpos on x-axis, F1 on y-axis, and speaker height (ht) color-coded. 

```{r}

```

Any generalizations or interesting observations?


## Saving plots

If you want to save your plots for future use, the command to use is `ggsave()`.
Here is an example:

```{r}
# By default, saves the last plot
ggsave("datv_ht1.pdf")

# You can designate a variable to a particular plot, then save:
p1 <- datv %>%
  ggplot(aes(x = ht, y = F0, col = sp_gender)) + 
  geom_point() 

# The full specifications of parameters
ggsave("datv_ht2.png", plot = p1, device = NULL, path = NULL,
  scale = 1, width = NA, height = NA, units = c("in", "cm", "mm"),
  dpi = 300, limitsize = TRUE)
```


# Visualizations from data summaries

Oftentimes, counts and mappings of raw data alone do not provide useful visualizations of experimental results. Recall from the last lab that we can create useful summary tables of a data set using the functions `summarise` and `group_by`.

What `summarise` does is to *apply a function* to a part of the dataset to create a new summary dataset. 

The general syntax: `summarise` takes multiple `new_column_name = function_to_be_applied_to_data(data_column)` entries in a list. 

Where these two verbs shine is in combination. Because `summarise` applies functions to columns in your *grouped data*, not just to the whole dataset!

So we can group by age or condition or whatever else we want and then carry out the same procedure, and all of a sudden we are doing something extremely useful!

```{r}
datv_means <- datv %>%
  group_by(sp_gender, wpos) %>%
  summarise(meanF0 = mean(F0))
datv_means
```

Let's plot the outputs of these summaries as follows:

```{r}
datv_means %>%
  ggplot(aes(x = wpos, y = meanF0, fill = sp_gender)) + 
  geom_bar(position=position_dodge(.9), stat = "identity")
```

What does the `position=` parameter specify? Figure it out by taking it out of the command and seeing what happens.



## Fine-tuning the summaries

Oftentimes, especially for bar plots, we would want to overlay error bars representing standard errors on top of mean values. We can easily create more sophisticated summary tables, using built-in functions:

```{r}
datv_means2 <- datv %>%
  group_by(sp_gender, wpos) %>%
  summarise(meanF0 = mean(F0),
            nF0 = n(),
            sdF0 = sd(F0), 
            seF0 = sdF0/sqrt(nF0))
datv_means2
```


Let's now create plots with error bars:

```{r}
datv_means2 %>%
  ggplot(aes(x = wpos, y = meanF0, fill = sp_gender)) + 
  geom_bar(position=position_dodge(.9), stat = "identity") +
  geom_errorbar(aes(ymin=meanF0-seF0, ymax=meanF0+seF0),
                width=.2,                    # Width of the error bars
                position=position_dodge(.9))
```



**Exercise 3** Generate a summary of datv which would enable you to examine whether and to what extent speaker height and speaker gender difference correlates with F1, F2, and F3 values, for each vowel. Include means and standard errors. Using the summary, generate a plot which maps height on the x-axis, F1 or F2 or F3 on the y-axis, and vowel types on separate facets/panels. You may also choose to color-code speaker gender.

```{r}

```

**Exercise 4** Examine datc. What could be some useful things to summarize and plot? Discuss among groups. Then implement a few visualizations.

```{r}

```

# Making things pretty

The possibilities are endless! Below are a few basic settings. For ideas on color combinations, check this out: [http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/) If you are a fan of Wes Anderson movies, you can also check this out: [https://github.com/karthik/wesanderson](https://github.com/karthik/wesanderson)

```{r}
# Basic
datv_means %>%
  ggplot(aes(x = wpos, y = meanF0, fill = sp_gender)) + 
  geom_bar(position=position_dodge(.9), stat = "identity") + 
  ggthemes::theme_few() + 
  ggthemes::scale_color_solarized() 

# Wes Anderson
install.packages("wesanderson")
library(wesanderson)

datv_means %>%
  ggplot(aes(x = wpos, y = meanF0, fill = sp_gender)) + 
  geom_bar(position=position_dodge(.9), stat = "identity") + 
  ggthemes::theme_few() + 
  scale_fill_manual(values = wes_palette("Royal1"))

```
