---
title: "R_tutorial3-uncommented"
author: "Sunwoo Jeong"
date: "12/1/2019"
output:
  pdf_document: default
  html_document: default
---

# The stop data

Let's begin by presenting your lab discussions from last time about our stop/vowel data from assignment 1.

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)

## The data
datc <- read.csv('asmt1-korean-stops.csv')
datg <- read.csv('asmt3-korean-ganong.csv')
```


# The Ganong data: overview

Recall the Ganong data. What kinds of data pre-processing might be helpful before we begin the visualizations?

```{r}
head(datg)
```

## Adding columns

First, due to counterbalancing, the core offline measure `response` does not refer to the same values across lab groups. Let's fix that by adding a column: `stop_response`; recall the `mutate()` function from our first lab; we'll use it with `ifelse()`

```{r}
summary(datg$word_choice0_eng)
summary(datg$word_choice1_eng)

datg2 <- datg %>%
  mutate(stop_response =
           ifelse((word_choice0_eng == "bama" | word_choice0_eng == "baneul") & response == "0", "b",
              ifelse((word_choice0_eng == "domang" | word_choice0_eng == "dongmil") & response == "0", "d",
                  ifelse((word_choice0_eng == "tomang" | word_choice0_eng == "tongmil") & response == "0", "t",
                     ifelse((word_choice0_eng == "pama" | word_choice0_eng == "paneul") & response == "0", "p",
                        ifelse((word_choice1_eng == "bama" | word_choice1_eng == "baneul") & response == "1", "b",
                           ifelse((word_choice1_eng == "domang" | word_choice1_eng == "dongmil") & response == "1", "d",
                             ifelse((word_choice1_eng == "tomang" | word_choice1_eng == "tongmil") & response == "1", "t",
                                ifelse((word_choice1_eng == "pama" | word_choice1_eng == "paneul") & response == "1", "p", "NA"
                                    )))))))))

datg2$stop_response <- as.factor(datg2$stop_response)
summary(datg2$stop_response)
```

*Exercise 1* It would also be useful to add a column which codes whether subjects made the aspirated or the lenis choice (glossing over stop types). Let's call this column `stop_response2`. Using `mutate()` and `ifelse()`, create this column; call in a summary of the resulting column

```{r}
datg2 <- datg2 %>%
  mutate(stop_response2 = FILL IN)

datg2$stop_response2 <- as.factor(datg2$stop_response2)
summary(datg2$stop_response2)
```

*Exercise 2* It would also be useful to add a column which can divide the data into two groups: ones in which the aspirated rendition constitutes a word, and ones in which the lenis rendition constitutes a word (glossing over stop types). Let's call this column `word_status`. Using `mutate()` and `ifelse()`, create this column; designate two levels: `len_word`, `asp_word` call in a summary of the resulting column.

```{r}
summary(datg2$base_word)

datg2 <- datg2 %>%
  mutate(word_status = 
           FILL IN)

datg2$word_status <- as.factor(datg2$word_status)
summary(datg2$word_status)
```

When visualizing the data, can you think of which prediction might benefit from the addition of this column?  


# Visualization I

Think of the best way to visualize the data to check if predictions 1 \& prediction 3 below hold. What should be plotted on the x-axis? What on the y-axis? (Recall the Ganong reading) What are the types of geom objects that would suit the data the most?

+ *Prediction 1* Other things being equal, higher F0 stimuli along the 7-step continua will elicit significantly more aspirated choices.
+ *Prediction 3* Other things being equal, there will be significantly more aspirated choices in the ambiguous F0 region for word:p/t-nonword:b/d type stimuli, and significantly more lenis choices in the ambiguous F0 region for word:b/d-nonword:p/t type stimuli.

To visualize the core offline measures (lenis vs. aspirated choice), we want to create a summary table of the count/proportion of lenis vs. aspirated responses as conditioned by the 7-step F0 stimuli. We will be subdividing the dataset into two groups based on word-status, and overlaying two plots from these two data frames.

First, recall our first R lab, and subdivide the data into 2 frames using `filter()`

We can use `summarise(n = n())` and `mutate(freq = n / sum(n))` functions to generate counts and percentages. Recall the joint use of `group_by()` and `summarise()` in creating summary tables.

We designate a new dataframe within an additional geom object to create overlays.

```{r}
datg2$F0 <- as.factor(datg2$F0)
summary(datg2$F0)

## Filter and create two data frames

datg_lenword <-
  datg2 %>%
  filter(word_status == "len_word")

datg_aspword <-
  FILL IN

## Create summary tables for each data frame

# lenis words

sum_lenword <- datg_lenword %>%
  group_by(F0, stop_response2) %>% 
  summarise(count=n()) %>% 
  mutate(perc=count/sum(count), n=sum(count))

sum_lenword

# aspirated words

sum_aspword <- datg_aspword %>%
  FILL IN

sum_aspword


## Visualization

sum_lenword %>%
  ggplot(aes(x = F0, y = perc, col = stop_response2)) + 
  geom_point() +
  geom_point(data = sum_aspword)

```


Do you see any problems? How do you want to change the plots? First, we may want to remove superfluous information, and plot only the lenis choices or only the aspirated choices. We can do that by filtering the summary tables. Let's also connect the dots as in the Ganong reading (what do these lines assume/signify?) -- we can do that by using the `group` command. I've also designated color-blind-friendly colors for each word status group, using the following reference: [Color palette from R cookbook](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/)

```{r}
## Filtering

sum_aspword2 <-
  sum_aspword %>%
  filter(stop_response2 == "asp")

sum_lenword2 <-
  FILL IN

## Visualization

sum_lenword2 %>%
  ggplot(aes(x = F0, y = perc)) + 
  geom_point(color = "#009E73") +
  geom_point(data = sum_aspword2, color = "#D55E00")

## Connecting points with lines

sum_lenword2 %>%
  ggplot(aes(x = F0, y = perc, group = stop_response2)) + 
  geom_point(color = "#009E73") + geom_line(color = "#009E73") +
  geom_point(data = sum_aspword2, color = "#D55E00") + geom_line(data = sum_aspword2, color = "#D55E00")
```

**Group Discussion**

Do the results confirm our predictions 1 \& 3? Why or why not? Discuss the theoretical implications of the results!

# Fine-tuning \& prettifying the plots

Anything else you want to add to the plots? Error bars would be helpful, right? 

## Adding error bars

For count data, we can add them using the formula below:

```{r}

## For the lenis words subset:

sum_lenword2$se = sqrt(sum_lenword2$perc*(1-sum_lenword2$perc)/sum_lenword2$n)
sum_lenword2$uci = sum_lenword2$perc + 1.96*sum_lenword2$se
sum_lenword2$lci = sum_lenword2$perc - 1.96*sum_lenword2$se

sum_lenword2

## Now do the same for the aspirated words subset:

FILL IN

```

Plots with error bars can be created as below (recall R lab 2);

```{r}

## Visualization with error bars

sum_lenword2 %>%
  ggplot(aes(x = F0, y = perc, group = stop_response2)) + 
  geom_point(color = "#009E73") + 
  geom_line(color = "#009E73") +
  geom_errorbar(color = "#009E73", 
                aes(ymin=lci, ymax=uci),
                width=.2,
                position=position_dodge(.9)) +
  geom_point(data = sum_aspword2, color = "#D55E00") + 
  geom_line(data = sum_aspword2, color = "#D55E00") +
  geom_errorbar(data = sum_aspword2, color = "#D55E00",
                aes(ymin=lci, ymax=uci),
                width=.2,
                position=position_dodge(.9))

```


# Visualization II

Now let us probe the data deeper to check if prediction 2 below holds. What should be plotted on the x-axis? What on the y-axis? Do we need additional geom objects like panels?

+ *Prediction 2* Other things being equal, stimuli created from aspirated base recordings will elicit significantly more aspirated choices than ones created from lenis base recordings.

Analogous to what we did when probing word status, we can create summary tables after subdividing data into stimuli created from lenis base and those created from aspirated base.

```{r}
## If you need, recall the column names again:
#head(datg)

## Filter and create two data frames

summary(datg$base_word)

datg_lenbase <-
  datg2 %>%
  filter(base_word == "baneul" | base_word == "bama" | base_word == "dongmil" | base_word == "domang")

datg_aspbase <-
  FILL IN

## Summarize

sum_lenbase <- 
  FILL IN

sum_lenbase

sum_aspbase <- 
  FILL IN

sum_aspbase

## Visualize

sum_lenbase %>%
  ggplot(aes(x = F0, y = perc, group = stop_response2)) + 
  geom_point(color = "#009E73") + geom_line(color = "#009E73") +
  geom_point(data = sum_aspbase, color = "#D55E00") + geom_line(data = sum_aspbase, color = "#D55E00")

## Add error bars

```

Discuss the results among groups. Do the results confirm prediction 2? (Recall the key assumptions needed to validate the predictions!) What are the theoretical implications?



# Visualization III

Let us now turn to visualizing the validity of prediction 4. What should be plotted on the x-axis? What on the y-axis? Do we need additional geom objects like panels?

+ *Prediction 4* Other things being equal, the effect of F0 on word choice (higher F0; more aspirated responses) will be greater when the speaker is female, and weaker when the speaker is male; conversely, the effect of VOT on word choice (aspirated base; more aspirated responses) will be greater when the speaker is male, and weaker when the speaker is female.

Given the nature of the hypothesis, let's start from the lenbase/aspbase subsets we created above, and further subdivide the data into 4 types.

```{r}
## If you need, recall the column names again:
#head(datg)
#head(datg_lenbase)

## Filter and create two sub data frames

datg_lenbasem <-
  datg_lenbase %>%
  filter(speaker == "M1") 

datg_lenbasef <-
  datg_lenbase %>%
  filter(speaker == "F1") 

datg_aspbasem <-
  datg_aspbase %>%
  filter(speaker == "M1") 

datg_aspbasef <-
  datg_aspbase %>%
  filter(speaker == "F1") 

sum_lenbasef <- datg_lenbasef %>%
  group_by(F0, stop_response2) %>% 
  summarise(count=n()) %>% 
  mutate(perc=count/sum(count), n=sum(count)) %>%
  filter(stop_response2 == "asp")

sum_aspbasef <- datg_aspbasef %>%
  group_by(F0, stop_response2) %>% 
  summarise(count=n()) %>% 
  mutate(perc=count/sum(count), n=sum(count)) %>%
  filter(stop_response2 == "asp")

sum_lenbasem <- datg_lenbasem %>%
  group_by(F0, stop_response2) %>% 
  summarise(count=n()) %>% 
  mutate(perc=count/sum(count), n=sum(count)) %>%
  filter(stop_response2 == "asp")

sum_aspbasem <- datg_aspbasem %>%
  group_by(F0, stop_response2) %>% 
  summarise(count=n()) %>% 
  mutate(perc=count/sum(count), n=sum(count)) %>%
  filter(stop_response2 == "asp")

## Visualize

sum_lenbasef %>%
  ggplot(aes(x = F0, y = perc, group = stop_response2)) + 
  geom_point(color = "#D55E00") + geom_line(color = "#D55E00", linetype = "dashed") +
  geom_point(data = sum_aspbasef, color = "#D55E00") + geom_line(data = sum_aspbasef, color = "#D55E00") +
  geom_point(data = sum_lenbasem, color = "#009E73") + geom_line(data = sum_lenbasem, color = "#009E73", linetype = "dashed") +
  geom_point(data = sum_aspbasem, color = "#009E73") + geom_line(data = sum_aspbasem, color = "#009E73")

## Add error bars

```

**Group Discussion**

What do the colors represent now? What do the dotted lines vs. solid lines represent? Do the results confirm prediction 4? Discuss among groups.


# Alternative ways of generating summaries

(Bonus -- no need to answer) Are there ways to generate a master summary table without subdividing the dataframes?

```{r}

count_sum <- datg2 %>%
  group_by(F0, word_status) %>%
  summarise(n_len = sum(stop_response2=="len"))

count_sum

```
